---
title: "Relational reasoning in L2 - data processing"
author: '[Guillermo Montero-Melis](https://www.mpi.nl/people/montero-melis-guillermo)'
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
output:
  html_document:
    depth: 2
    number_sections: yes
    theme: default
    toc: yes
---


Introduction
============

This script processes the raw data as provided by Petrus I and documents each
step. 

NB:

- I've previously saved all data files into the folder `data/data-petrus`.
- Because the data contains non-standard Swedish characters, all files have been
converted to UTF-8 encoding.


Setup workspace
===============


```{r setup, include=TRUE}
library(knitr)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```



Form/category task
==================

Load the data as provided by Petrus. 

```{r}
# d(ata) for f(orm) and c(ategory) task
d_fc <- read.csv("data-petrus/form-category-task_trial-level-data.csv", sep = ";", 
                 fileEncoding = "UTF-8", stringsAsFactors = FALSE)
head(d_fc)
```

Note there are unwanted columns (this happens often when converting Excel 
files into CSV). Remove them

```{r}
# We also rename some of the columns
d_fc <- d_fc %>%
  select(Subject, Task, Language = Lang., Trial, Score, ItemID = Item, Comment = notes)
kable(head(d_fc))
```


