---
title: "Relational reasoning in L2 - data processing"
author: '[Guillermo Montero-Melis](https://www.mpi.nl/people/montero-melis-guillermo)'
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
output:
  html_document:
    depth: 2
    number_sections: yes
    theme: default
    toc: yes
---


Introduction
============

This script processes the raw data as provided by Petrus I and documents each
step. 

NB:

- I've previously saved all data files into the folder `data/data-petrus`.
- Because the data contains non-standard Swedish characters, all files have been
converted to UTF-8 encoding.


Setup workspace
===============


```{r setup, include=TRUE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lme4)
library(GGally)
```



Form/category task
==================


Trial-level data
-----------------

Load the data as provided by Petrus. 

```{r}
# d(ata) for f(orm) and c(ategory) task
d_fc <- read.csv("data-petrus/form-category-task_trial-level-data.csv", sep = ";", 
                 fileEncoding = "UTF-8", stringsAsFactors = FALSE)
head(d_fc)
```

Note there are unwanted columns (this happens often when converting Excel 
files into CSV). Remove them:

```{r}
# We also rename some of the columns
d_fc <- d_fc %>%
  select(Subject, Task, Language = Lang., Trial, Score, ItemID = Item, 
         Comment = notes)
kable(head(d_fc))
```


### Sanity checks

Number of subjects (unique Subject ID's):
`r length(unique(d_fc$Subject))`

How many data points using different cross-tabulations:

```{r}
with(d_fc, table(Task, Subject, Language))
```

Looks like the data set is complete and balanced.

How many values are there in the table above that are not equal to 24?

```{r}
sum(with(d_fc, table(Task, Subject, Language)) != 24)
```


We can look at the same thing from the point of view of the items:

```{r}
with(d_fc, table(Task, ItemID, Language))
sum(with(d_fc, table(Task, ItemID, Language)) != 24)
```

Again, everything is balanced and there is no missing data.


Item data
---------

```{r}
# Load item data
fc_items <- read.csv("data-petrus/form-category-task_item-description.csv",
                     sep = ";", fileEncoding = "UTF-8", stringsAsFactors = FALSE)
head(fc_items)
```

```{r}
# Rename columns so they match the trial-level data
fc_items <- fc_items %>% rename(Task = uppgift, Language = språk, ItemID = item,
                                Descr = beskrivning)
kable(head(fc_items))
```

Again, some sanity checks:

```{r}
with(fc_items, table(Task, Language))
```

Looks good!


Now let's look at the actual descriptions of the items (`Descr` column).
How often does each description appear?

```{r}
fc_items %>% 
  group_by(Descr) %>% 
  summarise(NumberOfIdenticalItems = n())
```

Some items appear twice and some only appear once.
What happens is that some items are the same for both tasks in a given language,
but others are not

```{r}
fc_items %>% 
  group_by(Descr) %>% 
  summarise(NumberOfIdenticalItems = n()) %>%
  group_by(NumberOfIdenticalItems) %>%
  summarise(Freq = n())
```

If we rearrange the data into wide format, we can more easily compare the items
within a language but across tasks:

```{r}
# From long to wide, see
# http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/
fc_items_w <- fc_items %>%
  spread(Task, Descr)
# Create variable to compare the two columns
fc_items_w$SameDescr <- with(fc_items_w, form == kategori)
# Show all items that are *not* the same
kable(fc_items_w[! fc_items_w$SameDescr, ])
```

So the 9 items that are different contain the same words, but ordered differently.

